Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-26 AnalyticStructure_main.tex

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$ 
are susceptible to exhibit bound-state long-range behaviour \emph{only} when $\omega < |\eps_p|$, 
while they behave as unbound states for values of $\omega$ above the threshold levels.

This fact has interesting implication in the evaluation of LR quantities.
%
% In the next section we will discuss the implication of the existence of these
% two distinct regimes on the numerical convergence of the observables computed within the LR framework. 
% Before doing this we propose an analogous analysis with the aim to understand the localization 
% properties of the excitation-based description of the LR.
% 
% The localization properties of fluctuation states allows us to assess the reliability of the linear response functional
% computed using the given basis set. 
%
Indeed, writing the response 
density in terms of fluctuation states implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibit the locality properties of the potentials $ \op V$, $\op \Phi$, $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.
We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the fluctuation states are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. The quantities $\braket{\psi_p \op O| f_p^\Phi(\omega)}$, 
being scalr products of bound-state-like objects, may thus be evaluated with computational setup that
are similar to the ones usually employed for GS calculations.
This consideration proves why, when a given computer code employs established GS numerical techniques, 
\emph{it is much easier to converge LR quantities for $\omega$ below threshold}.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enable to express 
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ (see.e.g. \cite{baroni2008,linlinKPM}).

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions. In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
\emph{i.e.} inside the scalar products of 
\eqref{LinearResponseFunctDef2}. 
In this case $\ket{f_p^\Phi(\omega)}$ is a unbound, delocalized function; a numerical treatment tailored for localized
functions will \emph{never} be able to discretise such function with arbitrary accuracy.  
In this context, the bound state character of $\{\bra{\psi_p}\op O\}$ behave as a regulator:
\emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
This observation indicates that, above threshold, a computational setup 
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able (even implicitly) to express an
unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.
The non-vanishing spherical-wave behaviour in the long range of $\ket{f_p^\Phi(\omega)}$ is therefore responsible
of this convergence difficulties, especially for high frequencies.

@@

From the equations of motion for the excited states to the Casida equations...

>>>>

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$ 
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$, 
while they behave as unbound states for values of $\omega$ above the threshold levels.

This fact has interesting implication in the evaluation of LR quantities.
%
% In the next section we will discuss the implication of the existence of these
% two distinct regimes on the numerical convergence of the observables computed within the LR framework. 
% Before doing this we propose an analogous analysis with the aim to understand the localization 
% properties of the excitation-based description of the LR.
% 
% The localization properties of fluctuation states allows us to assess the reliability of the linear response functional
% computed using the given basis set. 
%
Indeed, writing the response 
density in terms of fluctuation states implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibit the locality properties of the potentials $ \op V$, $\op \Phi$, $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.
We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the fluctuation states are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. The quantities $\braket{\psi_p \op O| f_p^\Phi(\omega)}$, 
being scalr products of bound-state-like objects, may thus be evaluated with computational setup that
are similar to the ones usually employed for GS calculations.
This consideration proves why, when a given computer code employs established GS numerical techniques, 
\emph{it is much easier to converge LR quantities for $\omega$ below threshold}.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enable to express 
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions. In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
\emph{i.e.} inside the scalar products of 
\eqref{LinearResponseFunctDef2}. 
In this case $\ket{f_p^\Phi(\omega)}$ is a unbound, delocalized function; a numerical treatment tailored for localized
functions will \emph{never} be able to discretise such function with arbitrary accuracy.  
In this context, the bound state character of $\{\bra{\psi_p}\op O\}$ behave as a regulator:
\emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
This observation indicates that, above threshold, a computational setup 
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able (even implicitly) to express an
unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.
The non-vanishing spherical-wave behavior in the long range of $\ket{f_p^\Phi(\omega)}$ is therefore responsible
of this convergence difficulties, especially for high frequencies.

@@

% We show that Casida equations are equivalent to the operator eigenvalue equations \eqref{ExcitationOperatorsDef1} when an explicit basis $\{\ket{\eta}\}$ for the empty states subspace is used,
% so that $\op Q_0 = \sum_{\eta}\ketbra{\eta}{\eta}$. Operators $\op O$ satisfying the transverse condition $\op O = \op O_{\perp}$ (see eq. \eqref{RhopTransverseDef1}) can be represented as 
% follows
% \be\lb{OperatorTransBasisExpDef1}
% \op O = \sum_{p\eta}\left(
% \excite{p}{\eta}\trace{\decay{\eta}{p}\op O} + \decay{\eta}{p}\trace{\excite{p}{\eta}\op O}
% \right) \;,
% \ee
% where the \emph{transition operators} $\excite{p}{\eta} \equiv \ketbra{\eta}{\psi_p}$ and $\decay{\eta}{p} \equiv \ketbra{\psi_p}{\eta}$
% are normalized as follows
% \be
% \trace{\excite{p}{\eta}\excite{q}{\rho}} = \trace{\decay{\eta}{p}\decay{\rho}{q}} = 0 \;, \quad
% \trace{\excite{p}{\eta}\decay{\rho}{q}} = \delta_{pq}\delta_{\eta\rho} \;. \nonumber
% \ee
% Expanding equations \eqref{ExcitationOperatorsDef1} in the basis of transitions provides a matrix equation for the coefficients of the expansion. In particular from the left action equation
% one has
% \be
% \sum_{q\rho}\mat{\trace{\decay{\alpha}{p}\, \sop{\mathcal L}\, \excite{q}{\beta}} &
% \trace{\decay{\alpha}{p}\, \sop{\mathcal L}\, \decay{\beta}{q}} \\
% \trace{\excite{p}{\alpha}\, \sop{\mathcal L}\, \excite{q}{\beta}} &
% \trace{\excite{p}{\alpha}\, \sop{\mathcal L}\, \decay{\beta}{q}}}
% \mat{X_a^{p\eta} \\ Y_a^{p\eta}} = \Omega_a \mat{X_a^{p\eta} \\ Y_a^{p\eta}}
% \ee

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this scope we introduce an explicit basis $\{\ket{\eta}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_\eta X^a_{p\eta}\ket{\eta} \;, \qq
\bra{\chi^a_p} = \sum_\eta\bra{\eta}Y^a_{p\eta} \;.
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element $\ket{\rho}$ of the basis provides a linear systems of equations for the coefficients
$X^a_{p\eta}$ and $Y^a_{p\eta}$, namely
\begin{align}
 \sum_\eta(H_{0\rho\eta}-\eps_p\delta_{\rho\eta})X^a_{p\eta} + \bra{\rho}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{p\rho} \;, \nn \\
 -\sum_\eta (H_{0\eta\rho}-\eps_p\delta_{\rho\eta})Y^a_{p\eta} - \bra{\psi_p}\op V'[\op E_a]\ket{\rho}& = \Omega_a Y^a_{p\rho} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{\eta} \equiv \ketbra{\eta}{\psi_p}$ and $\decay{\eta}{p} \equiv \ketbra{\psi_p}{\eta}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{p\eta}\excite{p}{\eta}X^a_{p\eta}+\decay{\eta}{p}Y^a_{p\eta} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{\rho}\op V'[\op E_a]\ket{\psi_p} = \sum_{q\eta}\left(U[\decay{\rho}{p};\excite{q}{\eta}]X^a_{q\eta} + U[\decay{\rho}{p};\decay{\eta}{q}]Y^a_{q\eta}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{\rho} = \sum_{q\eta}\left(U[\excite{p}{\rho};\excite{q}{\eta}]X^a_{q\eta} + U[\excite{p}{\rho};\decay{\eta}{q}]Y^a_{q\eta}\right) \;, \nn
\end{align}
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{q\eta}\mat{F_{p\rho}^{q\eta} &  D_{p\rho}^{q\eta}  \\
- E_{p\rho}^{q\eta} & - G_{p\rho}^{q\eta} }
\mat{X^a_{q\eta} \\ Y^a_{q\eta}} = \Omega_a \mat{X^a_{p\rho} \\ Y^a_{p\rho}}
\ee
where
\begin{align}
& F_{p\rho}^{q\eta} = (H_{0\rho\eta}-\eps_p\delta_{\rho\eta})\delta_{pq} + U[\decay{\rho}{p};\excite{q}{\eta}] \;,\nn \\
& D_{p\rho}^{q\eta} = U[\decay{\rho}{p};\decay{\eta}{q}] \;,\nn \\
& E_{p\rho}^{q\eta} = U[\excite{p}{\rho};\excite{q}{\eta}] \;,\nn \\
& G_{p\rho}^{q\eta} = (H_{0\eta\rho}-\eps_p\delta_{\rho\eta})\delta_{pq} + U[\excite{p}{\rho};\decay{\eta}{q}] \;. \nn
\end{align}
Finally, exploiting the symmetry properties of both the unperturbed Hamiltonian and of the coupling kernel it follows that $F=G$ and $D=E$ so that equations
\eqref{ExcitationMatrixEq1} reduce to the usual formulation of the Casida equations in the basis of transitions.
